Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference

### 2018-08-15 reviewers_response.md

For now, this are just notes, we should make this a polished response.

## Open questions

1. Do we need PCA? I think that we don't and I have removed it, but if you feel strongly for including it, we can put it back.
2. Do we need to include all the convoluted preprocessing to go from GEO to a SingleCellExperiment? I actually have a SingleCellExperiment object saved on Github. I see arguments both for and against... it may actually be useful for people to see how to create a SCE from scratch.
3. Should we include spike-ins in the SingleCellExperiment object? I think so because one reviewer explicitly mentions it, but it makes it harder to carry out all the steps. For now, I am keeping them, showing with some EDA that are not very useful and removing them.
4. Discussion about choice of K. AIC/BIC don't seem to provide a good answer. What should we do?
5. Many of the parameters that we specify in RSEC are default. On one hand it would be nice to leave them so that we don't have to explain them (see comment from reviewer 2), but on the other hand it would be more reproducible if we decide to change the default values later... what should we do?

@@

> 3. In ZINB-WaVE, the authors specify the number of dimensions for the low-dimensional space (K) to be K=50. Could the authors add more details for the reader explaining why they picked K=50 and describe situations in which a user would want to specify a higher or lower K? In particular, it would be useful to discuss computational time in terms of number of genes and cells. Also, it would be useful to note that if you only wanted to use ZINB-WaVE to remove known covariates for normalization, you can use K=0.

The question of how to choose K is a very important one and a more comprehensive discussion on the role of K, both on accuracy and computational time, is in the original ZINB-WaVE paper. However, we agree that giving more guidance on the choice of K is a needed addition to this workflow. We have added a section, called "Choice of K", that discusses two functions, `zinbAIC` and `zinbBIC`, that can be used to decide the best value of `K` for a given dataset.

> Minor comments:
> 1. When selecting the top 1000 HVGs, why do the authors not take into account the overall mean-variance relationship and only select genes based on the variance?

This is mostly for illustration purposes, but we do find in our experience that naively selecting most variable genes gives a good set of informative genes, perhaps because it selects on/off genes. We have expanded this section to add a discussion about alternative strategies, which use the mean-variance relation observed in RNA-seq data.

@@

> It would be useful to put a link to the source code (or to the section where the link to the source code exists) near the top of the document.

We have modified the section "Package version" to include a link to the source code as well as instructions on how to install the packages needed by the workflow.

> I was confused a bit by "the first major bifurcation in the HBC lineage trajectory occurs prior to cell division". Can you be more specific about what you are referring to here by cell division, as without knowledge of the system, I'm not sure where the cell division you refer to should appear.

Thank you, we agree that the paragraph taken out of context is unclear. The first bifurcation identified by slingshot (occurring right after the clusters marked by DeltaHBC1 and DeltaHBC2 in Figure 2) produces two lineage trajectories: the first one (purple curve in Figure 2) generates sustentacular cells through a differentiation process characterized by the absence of cell cycle and division. To simplify, each HBC stem cell gives rise to one sustentacular cell. The other lineage trajectories (blue and orange curves in Figure 2) generate proliferative GBCs, which are characterized by cell cycle and division. These, in turn, generate olfactory sensory neurons and microvillous cells. We hope that this will be useful to better understand our system. We have updated the text to add this more detailed explanation.

@@

> On filtering for most variable genes, I understand this decision, and I also recommend it during workshops before making ordination plots. I know that students are not always certain why we care about variance (unsupervised). I like to mention that these are the genes where the "action" is. A side point, the log(x+1) is not variance stabilizing for RNA-seq counts in general. This filter can give higher priority to low count genes than to genes where there is interesting biological variability (though I do not doubt that the very high biological variance genes will be preserved). It might be useful to show a vsn::meanSdPlot() for the matrix log1p(assay(se))?

Thank you, this is a very good point. See also comment by Reviewer 1 about modeling the mean-variance relationship. We have expanded this section to add a discussion about why we care about the most variable genes and how different strategies that account for the structure of the data can be employed.

> "correcting for batch effects": What are batch effects? (Of course, I know what they are, but a reader may not, and you could cite some of the single cell literature here.)

We have moved our discussion on batch effects to a new section earlier in the workflow and we have expanded it to include an explanation of what we mean by "batch effects" in this specific example and more generally.

@@

> "Note that, in this case, the low-dimensional matrix W is not included in the computation of residuals to avoid the removal of the biological signal of interest.": I understood this sentence only on a second pass through. One problem is that you haven't defined W in the text yet (only in Figure 4). I would only reference the matrix W if you have defined it.

We agree that the reference to W taken out of context is hard to understand. In the revised version we introduce the matrix W in the previous paragraph making it more explicit what is its interpretation.

@@

> Figure 15 refers back to Figure 2 but does not use the same color scheme for the known cell types, so the reader cannot verify if you've recovered the lineages from the publication. It would be good therefore to have a legend for these figures (Fig 15 and following) indicating which cell types the colors refer to (this information is in the unlabeled table above, but should be included as a legend here).

Thanks for the suggestion, we agree that having a similar color scheme will help the reader make the connection between the results of Figure 15 and Figure 2. We have changed the colors to match as close as possible those of the published figure. Note that we do this right after the clustering step, so that all figures (from Figure 8 to Figure 17) have consistent colors.

> Can you briefly describe what a GAM is ahead of Figure 17?

We have added a sentence describing GAMs and a reference ahead of Figure 17.

### 2018-08-15 workflow/workflow.Rmd
Finish reviewers response

We remove the ERCC spike-in sequences and the CreER gene, as the latter corresponds to the estrogen receptor fused to Cre recombinase (Cre-ER), which is used to activate HBCs into differentiation following injection of tamoxifen (see [@Fletcher2017] for details).

DR: Consider keeping the spike-ins in the object to show the ability of SCE to use spike-ins... but we don't use them, so not sure how important it is.

```{r}
# Remove ERCC and CreER genes
cre <- E["CreER",]
ercc <- E[grep("^ERCC-", rownames(E)),]
E <- E[grep("^ERCC-", rownames(E), invert = TRUE), ]
E <- E[-which(rownames(E)=="CreER"), ]
dim(E)
```

@@

```{r}
# Extract QC metrics
qc <- as.matrix(protocolData(Cufflinks_eSet)@data)[,c(1:5, 10:18)]
qc <- cbind(qc, CreER = cre, ERCC_reads = colSums(ercc))

@@

Using the Bioconductor R package `scone` [@Cole235382], we remove low-quality cells according to the quality control filter implemented in the function `metric_sample_filter` and based on the following criteria (Figure \@ref(fig:scone)): (1) Filter out samples with low total number of reads or low alignment percentage and (2) filter out samples with a low detection rate for housekeeping genes. See the [scone vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/scone/inst/doc/sconeTutorial.html) for details on the filtering procedure.

@@

Metadata for the cells are stored in the slot `colData` from the `SummarizedExperiment` object. Cells were processed in 18 different runs of the Fluidigm C1 system: we refer to each run as a "batch", and we refer to the technical differences between each run as "batch effects". Note that in this particular case, the batch structure is essentially defined by the limitation of the C1 system to process up to 96 cells at a time and it is hence natural to identify each C1 run as one batch. Batch effects are ubiquitous in high-throughput genomics [@leek2010tackling] and single-cell RNA-seq is not an exception [@hicks2017missing]. In other contexts, depending on experimental design, batches may correspond to different library preparations, different labs in which the data is processed, or different times at which samples were collected.

@@

As with most dimensionality reduction methods, the user needs to specify the number of dimensions for the new low-dimensional space. Here, we use `K = 50` dimensions and adjust for batch effects via the matrix `X`. Note that if the users include more genes in the analysis, it may be preferable to reduce `K` to achieve a similar computational time.

@@

```{r boxplotNorm, fig.cap="ZINB-WaVE: Boxplots of normalized expression measures (deviance residuals), color-coded by batch."}
norm_order <- norm[, order(as.numeric(batch))]
col_order <- col_batch[batch[order(as.numeric(batch))]]
boxplot(norm_order, col = col_order, staplewex = 0, outline = 0,
        border = col_order, xaxt = "n", ylab="Expression measure")
abline(h=0)
```

@@

```{r mdsW, fig.cap="ZINB-WaVE: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by original published clustering."}
d <- dist(W)
fit <- cmdscale(d, eig = TRUE, k = 2)
plot(fit$points, col = col_clus[as.character(publishedClusters)], main = "",
     pch = 20, xlab = "Component 1", ylab = "Component 2")
legend(x = "topleft", legend = unique(names(col_clus)), cex = .5, 
       fill = unique(col_clus), title = "Sample")
```

@@

```{r rsec_50,eval=FALSE}
print(system.time(dataObj <- RSEC(dataObj, k0s = 4:15, alphas = c(0.1),
                                betas = 0.8, reduceMethod="zinbwave",
                                clusterFunction = "hierarchical01", minSizes=1,
                                ncores = NCORES, isCount=FALSE,
                                dendroReduce="zinbwave",
                                subsampleArgs = list(resamp.num=100,
                                           clusterFunction="kmeans",
                                          clusterArgs=list(nstart=10)),
                                verbose=TRUE,
                                consensusProportion = 0.7,
                                mergeMethod = "none", random.seed=424242,
                                consensusMinSize = 10)))
```

@@

The correspondance between the clusters we found here and the original clusters is as follows.

```{r tabagain}
table(data.frame(original = publishedClusters, ours = primaryClusterNamed(dataObj)))
```

@@

## Differential expression between clusters

DR: I think that it will be nice to showcase the getBestFeatures function here as well as our observational weights.

@@

Many different methods have been proposed for the analysis of scRNA-seq data and this paper presents only one possible workflow, using tools that we have developed in recent years. The use of the `SingleCellExperiment` class, which is the shared "language" among Bioconductor's scRNA-seq software packages, makes it extremely easy to change one of the step above to use a different approach. For instance, one could use `scPipe` [@tian2017scpipe] for gene expression quantification, `scater` [@McCarthy2017] for quality control and filtering, `scran` [@lun2016pooling] to robustly calculate global scaling factors for normalization, and `SC3` [@kiselev2017sc3] for clustering, with minimal to the code structure above. This makes the present workflow a truly modular approach, in which a user can use their favorite tool for each step with the guarantee that the output of the step will be readable by the next step.

An alternative normalization strategy is to compute gene and cell-specific scaling factors (e.g., `SCNorm` [@bacher2017scnorm]). Similarly, `monocle` [@trapnell2014dynamics] or `TSCAN` [@ji2016tscan] for cell lineage inference. Finally, the software suites `Seurat` [@seurat] (available on CRAN) and `scde` [@kharchenko2014bayesian] provide alternative pipelines for cell type discovery, differential expression, and gene set analysis. A comprehensive list of tools developed for the analysis of scRNA-seq data is available through at https://www.scrna-tools.org [@zappia2018exploring].

@@

DR: Not sure if we need this section. The previous future developments are now done (nice!). Perhaps here we should talk about scalability and HDF5? Kelly, is there anything you want to add here on DE along lineages?

>>>>

@@

```{r}
# Extract QC metrics
qc <- as.matrix(protocolData(Cufflinks_eSet)@data)[,c(1:5, 10:18)]

@@

There are three special sets of genes in this dataset: the CreER gene corresponds to the estrogen receptor fused to Cre recombinase (Cre-ER), which is used to activate HBCs into differentiation following injection of tamoxifen (see [@Fletcher2017] for details); the Sox2-eGFP transgenic was used for FACS sorting; the ERCC spike-in sequences are synthetic genes spiked-in at known concentrations and used as quality controls.

We can specify spike-in genes withing the `SingleCellExperiment` class. In this way, it is easier to retrieve their expression. In addition, some specific functions may make direct use of this information (see, e.g., the `computeSpikeFactors` function in the `scran` package).

In this case, we specify two different sets of spike-ins, the ERCC spike-ins and the transgenic genes (eGFP and Cre-ER). This is a slight abuse of notation, as the CreER and eGFP genes are not exactly spike-ins, but it is nonetheless useful to specify that this is not an endogenous gene.

```{r}
# Mark ERCC and CreER genes as spike-ins
isSpike(dataObj, "ERCC") <- grepl("^ERCC-", rownames(dataObj))
isSpike(dataObj, "Transgene") <- grepl("^CreER$", rownames(dataObj)) | 
                                  grepl("^eGFP$", rownames(dataObj)) 
```

We can access a specific set of spike-ins (or all of them) by using the `isSpike` function.

```{r}
table(isSpike(dataObj))
table(isSpike(dataObj, "Transgene"))
```

The total number of reads mapped to spike-ins is often a good quality-control measure. Hence, we include that into our `colData`.

```{r}
ercc <- counts(dataObj[isSpike(dataObj, "ERCC"),])
colData(dataObj)$ERCC_reads <- colSums(ercc)
```

## Exploratory data analysis

Cells were processed in 18 different runs of the Fluidigm C1 system: we refer to each run as a "batch", and we refer to the technical differences between each run as "batch effects". Note that in this particular case, the batch structure is essentially defined by the limitation of the C1 system to process up to 96 cells at a time and it is hence natural to identify each C1 run as one batch. Batch effects are ubiquitous in high-throughput genomics [@leek2010tackling] and single-cell RNA-seq is not an exception [@hicks2017missing]. In other contexts, depending on experimental design, batches may correspond to different library preparations, different labs in which the data is processed, or different times at which samples were collected.

Often, different batches differ in terms of data quality. In particular, we observed that the number of reads mapping to spike-ins can differ dramatically across batches.

```{r spikebatch, fig.cap="Distribution of reads mapped onto ERCC spike-ins per batch."}
boxplot(colData(dataObj)$ERCC_reads~batch, col = col_batch,
        outline = 0, ylab="Expression measure", las = 3)
```

Figure \@ref(fig:spikebatch) shows how for some batches no reads are mapped to ERCC spike-ins, presumably because the spike-ins were not included. Since the utility of the spike-ins is limited in this example, we remove the spike-ins from our object.

```{r}
dataObj <- dataObj[!isSpike(dataObj),]
dataObj <- clearSpikes(dataObj)
```

## Sample filtering

Using the Bioconductor package `scone` [@Cole235382], we remove low-quality cells according to the quality control filter implemented in the function `metric_sample_filter` and based on the following criteria (Figure \@ref(fig:scone)): (1) Filter out samples with low total number of reads or low alignment percentage and (2) filter out samples with a low detection rate for housekeeping genes. For each criterion, a sample is filtered out if it is below a user-defined hard threshold (which we leave at default value) or an adaptive threshold, defined in terms of number of standard deviations from the mean or median absolute deviation from the median.

The main parameter to set in this function is `zcut`, which represents the number of standard deviations from the mean and median absolute deviation from the median to use as threshold. We empirically found that using a threshold of three works well for typical datasets.

See the [scone vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/scone/inst/doc/sconeTutorial.html) for details on the filtering procedure.

@@

@@

The model can be thought of as a generalization of a factor analysis model, in which the biological signal of interest is assumed to be explained by a small number of "factors", represented in the model by the matrix `W` (Figure \@ref(fig:zinbschema)).

As with most dimensionality reduction methods, the user needs to specify the number of factors (`K`, the number of columns of `W`), which correspond to the dimensions for the new low-dimensional space. Here, we use `K = 50` dimensions and adjust for batch effects via the matrix `X`. Note that if the users include more genes in the analysis, it may be preferable to reduce `K` to achieve a similar computational time.

@@

```{r boxplotNorm, fig.cap="ZINB-WaVE: Boxplots of normalized expression measures (deviance residuals), color-coded by batch."}
batch <- colData(dataObj)$Batch
norm_order <- norm[, order(as.numeric(batch))]
col_order <- col_batch[batch[order(as.numeric(batch))]]
boxplot(norm_order, col = col_order, staplewex = 0, outline = 0,
        border = col_order, xaxt = "n", ylab="Expression measure")
abline(h=0)
```

@@

```{r mdsW, fig.cap="ZINB-WaVE: MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by original published clustering."}
d <- dist(W)
fit <- cmdscale(d, eig = TRUE, k = 2)
plot(fit$points, col = col_clus[as.character(colData(dataObj)$publishedClusters)],
     main = "", pch = 20, xlab = "Component 1", ylab = "Component 2")
legend(x = "topleft", legend = unique(names(col_clus)), cex = .5, 
       fill = unique(col_clus), title = "Sample")
```

@@

```{r rsec_50,eval=FALSE}
print(system.time(dataObj <- RSEC(dataObj, k0s = 4:15, alphas = c(0.1),
                                betas = 0.8, reduceMethod="zinbwave",
                                clusterFunction = "hierarchical01",
                                ncores = NCORES,
                                dendroReduce="zinbwave",
                                subsampleArgs = list(resamp.num=100,
                                           clusterFunction="kmeans",
                                          clusterArgs=list(nstart=10)),
                                consensusProportion = 0.7,
                                mergeMethod = "none", random.seed=424242,
                                consensusMinSize = 10)))
```

@@

As one can see from the code above, there are many parameters that a user can set in RSEC. Many of the values that we choose are the default values of the function, but we nonetheless specify them for clarity and reproducibility. Here, we briefly describethe role of the parameters that we set, but a more comprehensive explanation is provided in the function's man page (`?RSEC`) and in [@Risso2018].

- `k0s`: the values of `k` to start the sequential algorithm.
- `alphas`: similarity required in the clustering of the subsampling similarity matrix (larger alphas are less stringent).
- `betas`: similarity required to deem a cluster "stable" in sequential clustering (larger betas require greater similarity).
- `dendroReduce`: the dimensionality reduction to use for the creation of the dendrogram of the found clusters.
- `subsampleArgs`: a list of parameters for the subsample clustering; here we specified the base clustering function as k-means, the number of resamplings and the number of starting values for the k-means algorithm.
- `consensusProportion`: the proportion of times that two sets of samples should be together in order to be grouped into a cluster.
- `mergeMethod`: our consensus clustering may find too many small clusters. If that's the case a merging step can be employed to merge clusters that do not show enough differential expression. Here, we skip this step by setting the method as "none".
- `consensusMinSize`: minimum size required for a set of samples to be considered in a cluster; clusters smaller than this number will be removed and the corresponding samples will be set to "unclustered" (`-1`).

The correspondence between the clusters we found here and the original clusters is as follows.

```{r tabagain}
table(data.frame(original = colData(dataObj)$publishedClusters,
                 ours = primaryClusterNamed(dataObj)))
```

@@

@@

Many different methods have been proposed for the analysis of scRNA-seq data and this paper presents only one possible workflow, using tools that we have developed in recent years. The use of the `SingleCellExperiment` class, which is the shared "language" among Bioconductor's scRNA-seq software packages, makes it extremely easy to change one of the step above to use a different approach. For instance, one could use `scPipe` [@tian2017scpipe] for gene expression quantification, `scater` [@McCarthy2017] for quality control and filtering, `scran` [@lun2016pooling] to robustly calculate global scaling factors for normalization, and `SC3` [@kiselev2017sc3] for clustering, with minimal chenges to the code structure above. This makes the present workflow a truly modular approach, in which a user can employ their favorite tool for each step with the guarantee that the output of the step will be readable by the next step.

An alternative normalization strategy is to compute gene and cell-specific scaling factors (e.g., `SCNorm` [@bacher2017scnorm]). Similarly, `monocle` [@trapnell2014dynamics] or `TSCAN` [@ji2016tscan] are alternatives for cell lineage inference. Finally, the software suites `Seurat` [@seurat] (available on CRAN) and `scde` [@kharchenko2014bayesian] provide alternative pipelines for cell type discovery, differential expression, and gene set analysis. A comprehensive list of tools developed for the analysis of scRNA-seq data is available at https://www.scrna-tools.org [@zappia2018exploring].

@@

Thanks to the availability of droplet-based protocols, such as Drop-Seq and 10X Genomics Chromium, it is now becoming routine to generate datasets made of tens of thousands of cells. In fact, big collaborative projects such as the Human Cell Atlas [@regev2017science] and the BRAIN Initiative Cell Census Network [@ecker2017brain] will soon generate millions, if not billions, of cells. Hence, there is a growing need for scalable methods.

When datasets are too large to be fully stored in memory, HDF5 files [@folk2011overview] can be used to keep the data on disk and realize in memory only the subset of data needed at each step. We are in the process of updating the `clusterExperiment` and `zinbwave` packages to work with HDF5 files and we are developing more scalable methods to deal with massive datasets.
